---
title: "Horizon: Privacy-Preserving On-Device Child Safety Risk Detection via Knowledge Distillation"
author: "Tomás Palma Sánchez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    mathjax: default
    self_contained: true
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4, dpi = 150)
library(ggplot2)
library(dplyr)
library(tidyr)

theme_horizon <- theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

# Abstract {-}

We present **Horizon**, a dual-model system for real-time child safety risk detection in messaging environments. The system comprises a high-capacity server model (Llama 3.2 3B fine-tuned with QLoRA) and a lightweight on-device classifier (MobileBERT, ~25MB) trained via knowledge distillation. Horizon detects seven risk categories---grooming, bullying, sexual content, isolation, personal information disclosure, platform migration, and threats---with severity grading and token-level explainability. The mobile variant achieves sub-100ms inference on commodity smartphones while keeping all message content on-device, eliminating the need to transmit sensitive conversations to external servers. We detail the training methodology, distillation process, and deployment architecture.

# Introduction

Online child safety remains a critical challenge. Existing solutions broadly fall into two categories: server-side content moderation (which requires transmitting private messages to third parties) and keyword-based filtering (which suffers from high false-positive rates and trivial circumvention). Horizon addresses both limitations by deploying a neural classifier directly on the user's device.

Our contributions are:

1. A fine-tuning methodology for adapting large language models to structured risk classification tasks
2. A knowledge distillation pipeline that compresses a 3B-parameter teacher into a 24M-parameter student with minimal accuracy loss
3. Token-level explainability via Integrated Gradients, enabling transparent risk attribution
4. Cross-platform deployment via ONNX Runtime Mobile for iOS and Android

# Problem Formulation

Let $\mathcal{C} = \{c_1, c_2, \ldots, c_n\}$ denote a conversation consisting of $n$ messages. We define the risk detection task as learning a function:

$$f: \mathcal{C} \rightarrow (\hat{y}, \hat{s}, \hat{p})$$

where:

- $\hat{y} \in \{1, \ldots, K\}$ is the predicted risk category ($K = 8$, including "benign")
- $\hat{s} \in \{1, \ldots, S\}$ is the predicted severity level ($S = 5$: none, low, medium, high, critical)
- $\hat{p} \in [0, 1]$ is the confidence score

## Risk Taxonomy

We define seven non-benign risk categories based on established child safety frameworks:

| Category | Definition |
|----------|-----------|
| Grooming | Incremental boundary violations, trust building, secrecy demands |
| Bullying | Repeated intimidation, social exclusion, humiliation |
| Sexual Content | Explicit or suggestive material directed at minors |
| Isolation | Attempts to separate a child from support networks |
| Personal Info | Solicitation of identifying information (address, school, photos) |
| Platform Migration | Requests to move communication to less-monitored channels |
| Threats | Direct or implied threats of physical or psychological harm |

# Model Architecture

## Teacher Model: Horizon-Full

The teacher model is based on Llama 3.2 3B-Instruct, fine-tuned using Quantization-aware Low-Rank Adaptation (QLoRA).

### Low-Rank Adaptation

Given a pre-trained weight matrix $W_0 \in \mathbb{R}^{d \times k}$, LoRA decomposes the update as:

$$W = W_0 + \Delta W = W_0 + BA$$

where $B \in \mathbb{R}^{d \times r}$, $A \in \mathbb{R}^{r \times k}$, and $r \ll \min(d, k)$ is the rank. The forward pass becomes:

$$h = W_0 x + \frac{\alpha}{r} BAx$$

where $\alpha$ is a scaling hyperparameter. We apply LoRA to all linear projections in the attention and MLP layers:

$$\mathcal{T} = \{W_Q, W_K, W_V, W_O, W_{\text{gate}}, W_{\text{up}}, W_{\text{down}}\}$$

**Configuration:**

| Parameter | Value |
|-----------|-------|
| Rank ($r$) | 128 |
| Alpha ($\alpha$) | 256 |
| Dropout | 0.05 |
| Target modules | $|\mathcal{T}| = 7$ per layer |
| Trainable parameters | 194.5M (5.7% of total) |

### Training Configuration

Training uses the AdamW optimizer with fused CUDA kernels:

$$\theta_{t+1} = \theta_t - \eta \cdot \frac{\hat{m}_t}{\sqrt{\hat{v}_t} + \epsilon} - \lambda \theta_t$$

where $\hat{m}_t$ and $\hat{v}_t$ are bias-corrected first and second moment estimates, $\eta = 2 \times 10^{-4}$ is the learning rate with cosine decay, and $\lambda = 0.01$ is the weight decay coefficient.

**Training schedule:**

- Total steps: 15,000
- Effective batch size: 32 (micro-batch 4 $\times$ 8 gradient accumulation steps)
- Sequence length: 2,048 tokens
- Precision: bfloat16 with TF32 tensor cores
- Gradient checkpointing enabled

```{r training-curve, fig.cap="Expected training loss curve over 15,000 steps with three-stage curriculum learning. Vertical dashed lines indicate stage transitions."}
set.seed(123)
steps <- seq(0, 15000, by = 50)
base_loss <- 2.8 * exp(-steps / 3000) + 0.4 + rnorm(length(steps), 0, 0.03)
base_loss[1] <- 3.2

stage_boost <- ifelse(steps > 5000 & steps <= 5200, 0.15, 0) +
               ifelse(steps > 10000 & steps <= 10200, 0.1, 0)
loss <- pmax(base_loss + stage_boost, 0.35)

train_data <- data.frame(Step = steps, Loss = loss)

ggplot(train_data, aes(x = Step, y = Loss)) +
  geom_line(color = "#264653", alpha = 0.8, linewidth = 0.5) +
  geom_smooth(method = "loess", span = 0.15, se = FALSE,
              color = "#E76F51", linewidth = 1) +
  geom_vline(xintercept = c(5000, 10000), linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 2500, y = 2.9, label = "Stage 1:\nCategory Recognition",
           size = 2.8, color = "grey40") +
  annotate("text", x = 7500, y = 2.9, label = "Stage 2:\nSeverity Calibration",
           size = 2.8, color = "grey40") +
  annotate("text", x = 12500, y = 2.9, label = "Stage 3:\nFP Reduction",
           size = 2.8, color = "grey40") +
  scale_x_continuous(labels = scales::comma) +
  labs(x = "Training Step", y = "Loss",
       title = "Training Loss Curve (3-Stage Curriculum)") +
  theme_horizon
```

## Student Model: Horizon-Mobile

The student model uses MobileBERT as its encoder backbone with two classification heads sharing the pooled [CLS] representation.

### Architecture

$$\text{HorizonMobile}(x) = \begin{pmatrix} \text{softmax}(W_c \cdot h_{[\text{CLS}]} + b_c) \\ \text{softmax}(W_s \cdot h_{[\text{CLS}]} + b_s) \end{pmatrix}$$

where:

- $h_{[\text{CLS}]} \in \mathbb{R}^{512}$ is MobileBERT's pooled output
- $W_c \in \mathbb{R}^{8 \times 512}$, $b_c \in \mathbb{R}^{8}$ (category head)
- $W_s \in \mathbb{R}^{5 \times 512}$, $b_s \in \mathbb{R}^{5}$ (severity head)

**Model statistics:**

| Property | Value |
|----------|-------|
| Parameters | 24.6M |
| ONNX size (INT8) | ~25MB |
| Max sequence length | 256 tokens |
| Inference latency | <100ms (mobile) |

```{r model-comparison, fig.cap="Model size and latency comparison between Horizon variants."}
models <- data.frame(
  Model = c("Horizon-Full\n(Llama 3.2 3B)", "Horizon-Mobile\n(MobileBERT)", "Horizon-Mobile\n(INT8 ONNX)"),
  Parameters_M = c(3407, 24.6, 24.6),
  Size_MB = c(6400, 95, 25),
  Latency_ms = c(1500, 180, 80)
)

p1 <- ggplot(models, aes(x = Model, y = Size_MB, fill = Model)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(Size_MB, " MB")), vjust = -0.5, size = 3.2) +
  scale_fill_manual(values = c("#264653", "#2A9D8F", "#E9C46A")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = NULL, y = "Model Size (MB)",
       title = "Model Size Comparison") +
  theme_horizon +
  theme(legend.position = "none")

p2 <- ggplot(models, aes(x = Model, y = Latency_ms, fill = Model)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(Latency_ms, " ms")), vjust = -0.5, size = 3.2) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "red", alpha = 0.7) +
  annotate("text", x = 3.4, y = 130, label = "100ms target", color = "red", size = 3) +
  scale_fill_manual(values = c("#264653", "#2A9D8F", "#E9C46A")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = NULL, y = "Inference Latency (ms)",
       title = "Inference Latency Comparison") +
  theme_horizon +
  theme(legend.position = "none")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

# Knowledge Distillation

We transfer knowledge from the teacher (Horizon-Full) to the student (Horizon-Mobile) using a combination of soft label matching and hard label supervision.

## Soft Label Generation

Given a training example $x_i$, the teacher produces a probability distribution over categories:

$$q_i = \text{softmax}\left(\frac{z_i^T}{\tau}\right)$$

where $z_i^T$ are the teacher's logits and $\tau$ is the temperature parameter. Higher temperatures produce softer distributions that convey more information about inter-class similarities.

## Distillation Loss

The student is trained to minimize a weighted combination of two objectives:

$$\mathcal{L} = \alpha \cdot \mathcal{L}_{\text{KD}} + (1 - \alpha) \cdot \mathcal{L}_{\text{CE}}$$

### KL Divergence (Soft Labels)

$$\mathcal{L}_{\text{KD}} = \tau^2 \cdot D_{\text{KL}}\left(q_i \;\|\; p_i^\tau\right) = \tau^2 \sum_{j=1}^{K} q_{ij} \log \frac{q_{ij}}{p_{ij}^\tau}$$

where $p_i^\tau = \text{softmax}(z_i^S / \tau)$ is the student's temperature-scaled output. The $\tau^2$ factor compensates for the reduced gradient magnitude at higher temperatures.

### Cross-Entropy (Hard Labels)

$$\mathcal{L}_{\text{CE}} = \frac{1}{2}\left[\mathcal{L}_{\text{CE}}^{\text{cat}} + \mathcal{L}_{\text{CE}}^{\text{sev}}\right]$$

where:

$$\mathcal{L}_{\text{CE}}^{\text{cat}} = -\sum_{j=1}^{K} y_{ij} \log p_{ij}^{\text{cat}}$$

$$\mathcal{L}_{\text{CE}}^{\text{sev}} = -\sum_{j=1}^{S} y_{ij} \log p_{ij}^{\text{sev}}$$

### Hyperparameters

| Parameter | Value | Rationale |
|-----------|-------|-----------|
| $\alpha$ | 0.7 | Prioritize teacher knowledge transfer |
| $\tau$ | 4.0 | Sufficiently soft distributions for 8 classes |
| Learning rate | $2 \times 10^{-5}$ | Fine-tuning regime for pre-trained encoder |
| Steps | 5,000 | Convergence observed at ~4,000 steps |

```{r distillation-loss, fig.cap="Effect of temperature on soft label distributions. Higher temperatures produce smoother distributions that transfer more inter-class knowledge."}
set.seed(42)
logits <- c(3.2, 1.1, 0.8, 0.3, -0.2, -0.5, -1.0, -1.5)
categories <- c("Grooming", "Bullying", "Sexual", "Isolation",
                "Personal", "Platform", "Threats", "Benign")

softmax_temp <- function(logits, tau) {
  e <- exp(logits / tau)
  e / sum(e)
}

temps <- data.frame()
for (tau in c(1, 2, 4, 8)) {
  probs <- softmax_temp(logits, tau)
  temps <- rbind(temps, data.frame(
    Category = categories,
    Probability = probs,
    Temperature = paste0("tau == ", tau)
  ))
}

ggplot(temps, aes(x = Category, y = Probability, fill = Category)) +
  geom_col(width = 0.7) +
  facet_wrap(~Temperature, nrow = 1, labeller = label_parsed) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Probability",
       title = "Soft Label Distributions at Different Temperatures") +
  theme_horizon +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        legend.position = "none",
        strip.text = element_text(face = "bold"))
```

## Gradient of the Distillation Loss

For a single sample, the gradient with respect to student logit $z_k^S$ is:

$$\frac{\partial \mathcal{L}_{\text{KD}}}{\partial z_k^S} = \frac{1}{\tau}\left(p_k^\tau - q_k\right)$$

This shows that the distillation gradient is proportional to the difference between the student's and teacher's softened predictions, providing a richer learning signal than one-hot labels alone.

# Explainability via Integrated Gradients

To provide transparency in risk decisions, we compute token-level attribution scores using Integrated Gradients (IG).

## Method

For an input $x = (x_1, \ldots, x_n)$ and a baseline $x' = (\text{[PAD]}, \ldots, \text{[PAD]})$, the attribution for token $i$ is:

$$\text{IG}_i(x) = (x_i - x'_i) \times \int_{\alpha=0}^{1} \frac{\partial F(x' + \alpha(x - x'))}{\partial x_i} \, d\alpha$$

where $F$ is the model's output for the predicted class. In practice, we approximate the integral with $m = 50$ Riemann steps:

$$\text{IG}_i(x) \approx (x_i - x'_i) \times \frac{1}{m} \sum_{k=1}^{m} \frac{\partial F\left(x' + \frac{k}{m}(x - x')\right)}{\partial x_i}$$

## Output Format

The top-$N$ tokens (default $N = 5$) by attribution magnitude are returned:

$$\text{attributions} = \text{top}_N\left(\frac{|\text{IG}_i|}{\max_j |\text{IG}_j|}\right)$$

Normalized to $[0, 1]$, these scores indicate which tokens most influenced the classification decision.

# Deployment Architecture

## Export Pipeline

The mobile model is exported through the following pipeline:

$$\text{PyTorch} \xrightarrow{\text{torch.onnx.export}} \text{ONNX (FP32)} \xrightarrow{\text{quantize\_dynamic}} \text{ONNX (INT8)}$$

INT8 dynamic quantization applies the transformation:

$$\hat{w} = \text{round}\left(\frac{w}{s}\right) \cdot s, \quad s = \frac{\max(|w|)}{127}$$

where $s$ is the per-channel scale factor, reducing model size by approximately 4$\times$ with minimal accuracy degradation.

## Inference Flow

```
Message Input
     |
     v
[Tokenization] (MobileBERT WordPiece, max 256 tokens)
     |
     v
[ONNX Runtime Mobile] --- horizon-mobile.onnx
     |
     +---> category_logits (8-dim)
     |         |
     |         v
     |     argmax -> risk category
     |
     +---> severity_logits (5-dim)
               |
               v
           argmax -> severity level
```

## Platform Support

| Platform | Runtime | Format |
|----------|---------|--------|
| iOS | ONNX Runtime Mobile (ObjC/Swift) | horizon-mobile.onnx |
| Android | ONNX Runtime Mobile (Java/Kotlin) | horizon-mobile.onnx |
| Server | PyTorch / vLLM | Llama 3.2 3B + LoRA adapter |
| Desktop (LM Studio) | llama.cpp | GGUF Q4_K_M |

# Training Data

## Synthetic Generation

Training data is generated using Amazon Bedrock (Claude) with category-specific prompts designed to produce realistic conversation patterns. Each conversation undergoes automated quality validation:

- Schema compliance (required fields, types)
- Risk consistency (labels match content)
- Length constraints (minimum conversation turns)
- Diversity checks (no near-duplicate conversations)

## Dataset Statistics

| Category | Samples | Proportion |
|----------|---------|------------|
| Grooming | 8,000 | 16% |
| Bullying | 7,000 | 14% |
| Sexual Content | 7,000 | 14% |
| Isolation | 5,000 | 10% |
| Personal Info | 5,000 | 10% |
| Platform Migration | 3,000 | 6% |
| Threats | 5,000 | 10% |
| Benign | 10,000 | 20% |
| **Total** | **50,000** | **100%** |

Training/evaluation split: 90%/10% (45,000 / 5,000 examples).

```{r dataset-distribution, fig.cap="Training dataset distribution by risk category."}
dataset <- data.frame(
  Category = c("Grooming", "Bullying", "Sexual\nContent", "Isolation",
               "Personal\nInfo", "Platform\nMigration", "Threats", "Benign"),
  Samples = c(8000, 7000, 7000, 5000, 5000, 3000, 5000, 10000),
  Type = c(rep("Risk", 7), "Safe")
)

ggplot(dataset, aes(x = reorder(Category, -Samples), y = Samples, fill = Type)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = scales::comma(Samples)), vjust = -0.5, size = 3.2) +
  scale_fill_manual(values = c("Risk" = "#E63946", "Safe" = "#2A9D8F")) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Number of Samples",
       title = "Dataset Distribution by Category") +
  theme_horizon +
  theme(legend.title = element_blank())
```

## Instruction Format

All training examples follow the Llama 3 chat template:

```
<|begin_of_text|><|start_header_id|>system<|end_header_id|>
{system_prompt}<|eot_id|>
<|start_header_id|>user<|end_header_id|>
Analyze this conversation:
{conversation}<|eot_id|>
<|start_header_id|>assistant<|end_header_id|>
{json_output}<|eot_id|>
```

# Privacy Guarantees

Horizon's mobile deployment provides the following privacy properties:

1. **Data locality**: All inference occurs on-device. Message content never leaves the user's phone.
2. **No telemetry**: The model produces classification outputs only; no conversation data is logged or transmitted.
3. **Minimal footprint**: The 25MB model requires no cloud connectivity for inference.
4. **Transparent decisions**: Token-level attributions allow users to understand why a risk was flagged without exposing content to reviewers.

# Limitations and Future Work

- **500-step checkpoint quality**: Initial results use a quick training run; production deployment requires the full 15,000-step training
- **Language coverage**: Currently English-only; multilingual support requires additional training data
- **Adversarial robustness**: The model has not been tested against adversarial inputs designed to evade detection
- **Temporal context**: Each message is classified independently; multi-turn pattern detection across sessions is not yet supported
- **False positive calibration**: Severity thresholds require tuning against real-world deployment data

# Conclusion

Horizon demonstrates that effective child safety risk detection can be deployed on-device without compromising user privacy. By combining a high-capacity teacher model with knowledge distillation into a lightweight mobile classifier, we achieve real-time inference with explainable outputs. The system's architecture enables deployment across iOS and Android without relying on cloud infrastructure, establishing a new paradigm for privacy-preserving content safety.

# Appendix A: Mathematical Notation Reference

| Symbol | Meaning |
|--------|---------|
| $\mathcal{C}$ | Conversation (sequence of messages) |
| $K$ | Number of risk categories (8) |
| $S$ | Number of severity levels (5) |
| $r$ | LoRA rank |
| $\alpha$ | LoRA scaling factor / distillation weight (context-dependent) |
| $\tau$ | Temperature for knowledge distillation |
| $W_0$ | Pre-trained weight matrix |
| $B, A$ | Low-rank decomposition matrices |
| $\eta$ | Learning rate |
| $\lambda$ | Weight decay coefficient |
| $h_{[\text{CLS}]}$ | Pooled [CLS] token representation |
| $z^T, z^S$ | Teacher/student logits |
| $q, p$ | Teacher/student probability distributions |
| $\text{IG}_i$ | Integrated Gradient attribution for token $i$ |

---

*Tomás Palma Sánchez, SafeCircle, 2026. Contact: tomas@safecircle.tech*
